---
title: "Amphibians in the Sierras: Task 1"
author: "Kacie Ring"
date: "1/19/2022"
output:
  html_document:
    theme: united
    highlight: tango

---

## Introduction 
        
        The objective of this assignment is to obtain a large meta dataframe and manipulate it to observe patterns in the collected empirical data. The data being analyzed in this assignment is a subset containg the observations of amphibian species from a larger data frame containing observations of the community composition of ponds and lakes in the Sierra Nevadas. Within the subset of the amphbian data there are 8,150 observations of the adundance of four life stages of 10 species across 7 years. Rana muscosa populations dynamics was of interest due to their enlistment as an endangered species following the introduction of invasive trout in lakes and ponds in the high sierras. The main questions being answered with the following data manipulation are: 1) How has the abundance of Rana Muscosa in ponds and lakes changed from 1995-2002 and 2) Which specific lakes have the highest abundance of Rana Muscosa adults and subadults. To acheive task 1, a subest of data only containing Rana Muscosa was created, the abundance data was then organized by life stage (tadpole, subadult, adult) and year. A stacked baoxplot was generated to observe the change in Rana muscosa abundances from 1995-2002. For task 2, the same subset of only Rana Muscosa was used and counts of subadult and adults were combined and grouped by Lake ID to identify which specific lakes had the highest species count. 
        
## Data Analysis 

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

#load in required packages 
library(readr)
library(tidyverse)
library(here)
library(patchwork)
library(ggplot2)
library(cowplot)
library(lubridate)

#read in dataframe 
frog <- read_csv(here('data', 'sierra_amphibians.csv'),
                     show_col_types = FALSE)
```


### Task 1 

**Objective**: A graph of total mountain yellow-legged frog (Rana muscosa) count each year across all water bodies.

**Variables of interest** in data frame and associated task:

  * *amphibian number* total by
  * *amphibian_species* -> RAMU (subset data to only include Ramu) group by 
  * *survey_date* ->  convert survey_data to year 
  * *amphibian_life_stage* -> remove Eggmass
  * *amphibian_location* -> water bodies.. exclude fringe habitat? No grouping 


#### Data wrangling 

```{r task1 -data wrangle}
 
#step one: subset RAMU and filter out Eggmass life stage 
rana <- frog %>%
  filter(amphibian_species == 'RAMU' & amphibian_life_stage != 'EggMass') %>%
  select("Count" = "amphibian_number", 
         "Date" = "survey_date", 
         "LifeStage" = "amphibian_life_stage", 
         "Habitat" =  "amphibian_location")

#step two: convert date 
rana <- rana %>%   
  mutate(date = lubridate::mdy(rana$Date))#mdy format 

rana <- rana %>%  
  mutate(year = lubridate::year(rana$date)) #pull year add to new column 


```

#### Create boxplot

```{r task 1 - create boxplot}

#create a stacked boxplot
task_1 <- ggplot(data = rana, 
                aes(x=as.factor(year), y=Count))  +
                geom_col(aes(fill= LifeStage), width = 1) + 
                scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73"), 
                       name="Life Stage", #legend label 
                       breaks=c("Adult", "SubAdult", "Tadpole"),
                       labels=c("Adult", "Sub Adult", "Tadpole")) + #legend names 
                      xlab("Year") + 
                      theme_minimal()  + 
            labs(title = "Mountain yellow-legged frog count by year") +
            theme(
            plot.title = element_text(hjust = 0.5, size = 10)) +  # Center title position and size
            theme(axis.text.x = element_text(angle = 90)) #flip axis labels 90 degress


 
```

### Task 2 

A column graph containing total counts of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts

Variables of interest in data frame and associated task: 

  * *Lake_ID* -> Top 5 lakes of 
  * *Life Stage* -> Adult and sub-adult populations with*
  * *Count* -> highest counts

#### Data wrangling 

```{r task 2 - data wrangle}

#step 1 data wrangling 

#create subset with variables of interest
top_5 <- frog %>%
  filter(amphibian_species == 'RAMU' & 
           amphibian_life_stage != 'EggMass' & 
           amphibian_life_stage !='Tadpole') %>% #select RAMU and remove Eggmass/Lifestages 
  select("lake_id", 
         "Count" = "amphibian_number", 
         "LifeStage" = "amphibian_life_stage") #rename columns 

#rename lake_id, group by lake_id, get total counts for lake_is(adult+sub_adult), reorder 
top_5 <- top_5 %>%
    mutate(lake_id = paste('Lake', lake_id, sep = ' ')) %>% #create Lake ids 
    group_by(lake_id) %>% #group by lake id 
    summarize(total = sum(Count)) %>% #combine adult and sub adult counts 
    slice_max(order_by = total, n = 5) %>% #top 5 
    mutate(lake_id = fct_reorder(lake_id, total)) #reorder lake ID by total count 
 

```


#### Create a column graph 

Graph layout

    * x-axis = lake id (categorical)
    * y-axis = yollow-legged frogs counts (continuous)
    

```{r}

#make column graph 

task_2 <- ggplot(data = top_5, aes(x = lake_id, y = total)) +
            geom_col(fill = 'cyan3', width = .75) +
            xlab("Lake ID") + 
            ylab("Count") +
            theme_minimal() + 
            labs(title = 'Mountain yellow-legged frog count by Lake') +
            theme(
            plot.title = element_text(hjust = 0.5, size = 10)) +  # Center title position and size
            theme(axis.text.x = element_text(angle = 90))
           


```

#### Combine graphs 

```{r, fig.align='center', fig.width= 9,fig.cap="Figure 1: Total mountain yellow-legged frog (Rana muscosa) counts A) for tadpoles, subadults, and adults each year across all water bodies and B) for adults and subadults among Lake IDs with highest frog counts", warning=FALSE}
#combine plots 
plot_grid(task_1, task_2, 
          labels = c('A', 'B'), #label figures A & B 
          label_size = 10, #Adjust label size 
          align = 'hv', #lign figures 
          rel_widths = c(3,2.5)) #make Fig A wider than Fig B 

```

## Data Reference 

Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project:         
    Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2.      Environmental Data Initiative. 
    https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3



