---
title: "World data"
description: The purpose of this analysis is to determine what environmental variables determine mean precipitation levels by country.|
output: 
  distill::distill_article:
    code_folding: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
 
# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(palmerpenguins)
library(patchwork)
 
# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs


world <- read.csv("world_env_vars.csv")


```

### Introduction 

The purpose of this analysis is to determine what environmental variables determine mean precipitation levels by country. We utilized the "Environmental variables for world countries" available on kaggle. 


### Statistical methods 

Data was organized by descending mean water precipitation and subsetted with into highest and lowest 20 precipitation mean values. The top 20 were categorized as wet countries and the lowest 20 countries were categorized as dry countries. Variables measuring rain were removed from the dataset to avoid any sample clustering based on rain values. Data was scaled and stored in a distance matrix. 


**Data Cleaning** : Categorize wet and dry countries. Remove precipitation variables and scale data. 


```{r}

world_wet <- world %>%
  drop_na() %>% 
  arrange(desc(rain_mean_annual)) %>%
  filter(rain_mean_annual > 2282.339) %>% #20 highest annual rain 
  mutate(Conditon = "Wet")

world_dry <- world %>%
  drop_na() %>% 
  arrange(desc(rain_mean_annual)) %>%
  filter(rain_mean_annual < 229.21234) %>%
   mutate(Conditon = "Dry")
  #20 lowest annual rain 
 
world_clean <- full_join(world_wet, world_dry) #rejoin data subsets 


world_pca <- world_clean %>%
  select(2:28) %>%  #drop country names 
  select(-c(3:4, 7:15, 19, 20, 22, 25)) %>%
  scale() %>% #scale 
  prcomp() #create pc matrix 

```

### Create biplot

```{r, fig.align='center', fig.cap='Figure 1: PCA plot showing the environmental variables that may affect precipitation levels in wet and dry countries.'}

autoplot(world_pca,
     	data = world_clean,
     	loadings = TRUE,
     	colour = 'Conditon', #wet and dry condition 
     	loadings.label = TRUE,
     	loadings.label.size = 2.5,
     	loadings.colour = "grey",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.4,
     	loadings.label.hjust = .05
     	)  +
  labs(main = "Biplot of wet and dry countries using environmental variables") + 
  scale_color_manual(values = c("darkgoldenrod2","aquamarine4")) +
  scale_fill_manual(values = c("darkgoldenrod2","aquamarine4")) +
  theme_minimal()
 

```

### Screeplot

```{r, fig.align='center', fig.cap='Figure 2: Screeplot showing additional principal component percentages from Figure 1'}

sd_vec <- world_pca$sdev #pulls standard deviation (we want the variance tho)
var_vec <- sd_vec^2 ### standard deviation is sqrt of variance!
pc_names <- colnames(world_pca$rotation) #make a vector of pc names
 
pct_expl_df <- data.frame(v = var_vec, 
                          pct_v = var_vec / sum(var_vec), 
                          #each individual value divided by the total 
                          pc = fct_inorder(pc_names)) %>%
                          #
  mutate(pct_lbl = paste0(round(pct_v*100, 1), '%')) #create a percentage 
 
ggplot(pct_expl_df, aes(x = pc, y = v)) +
  geom_col() +
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) +
  labs(x = 'Principal component', y = 'Variance explained', main = "Screeplot") + 
  theme_minimal()
```



### Conclusion 

Our principle component analysis shows that certain environmental variables are associated with wet and dry countries. Cloudiness and canopy cover are strongly associated with wet countries whereas cropland, wind, and acccessibility to cities are associated with dry countries. This analysis shows that certain human associated land use alterations such as deforestation, conversion to agriculture,and urban development may reduce mean precipitation levels. 






